### 2017-07-03 reasoning.tex
Edits.

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference
may include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to the the inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does
extend naturally to isolation-constrained interference ($\R$)
because $R^{*}$ modulo $\I$ is not same as $\R^{*}$; the former is a
transitive relation constrained by $\I$, whereas the latter is the
transitive closure of a $\I$-constrained relation. We therefore
introduce a side-condition on $\I$ that restores the equality. The
condition requires $\I$ to allow an intereference $R^{*}(\stg,\stg'')$,
for two database states $\stg$ and $\stg''$, only if it also allows
interference for every prefix of $R^{*}(\stg,\stg'')$. In other words,
if $\I$ disallows interference from $\stg$ to $\stg'$, then an
$R$-step from $\stg'$ to $\stg''$ shouldn't make the interference from
$\stg$ to $\stg''$ valid. We call this the stability condition on
$\I$, defined as below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if is
invalid, then an interference has already modified a record, and no
further interference will restore the original record. This is because
records bear transaction ids, and the transaction that wrote the
original record has long committed.

@@

% \begin{definition}[\bfseries Step-indexed reflexive transitive closure]
% For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
% \mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
% the smallest relation satisfying the following
% properties:
% \begin{itemize}
% \item $\forall (x:A).\, R^0 (x,x)$
% \item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
% R^{n+1}(x,z)$
% \end{itemize}
% \end{definition}

% \begin{definition}[\bfseries Interleaved step and multi-step relations]
% An interleaved step relation interleaves thread-local reductions with
% interference from concurrent threads captured as the rely relation
% ($R$).  It is defined thus:\vspace*{-10pt}

% \begin{smathpar}
% \begin{array}{lcl}
% \I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
%   (c,\E) \stepsto (c',\E') \\
%   &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
% \end{array}
% \end{smathpar}

% \noindent The interleaved step relation for transaction bound expressions
% ($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
% An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
% reflexive transitive closure of the interleaved step relation.
% \end{definition}

% \begin{definition}[\bfseries Semantics of the RG judgment]
% \label{def:rg-semantics}
% The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
% in terms of the interleaved step relation thus:\vspace*{-10pt}

% \begin{smathpar}
% \begin{array}{l}
% \hspace*{-0.3in}
% \rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
%   \wedge \mathbb{I}(\E) \\
% \hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
%     \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
% \hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
% \end{array}
% \end{smathpar}

% \noindent The first conjunct in the consequent is called the \emph{Hoare
% consequent} since it ascribes Hoare triple semantics to an RG sextuple.
% The second conjunct, called the \emph{guarantee consequent}, uses the
% $\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}

% \begin{smathpar}
% \begin{array}{l}
% \texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
% \hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
%   (c'',\E'') \Rightarrow G(\E',\E'')\\
% \end{array}
% \end{smathpar}

% \noindent The guarantee consequent requires $G$ to capture the trace effect of
% every small-step of $c$, where the reduction can be interleaved by the
% interference ($R$) from concurrent threads. The semantics of the RG
% sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
% similarly. Expressions, unlike commands, evaluate to a value $v$, and
% the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
% Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
% \end{definition}

% Note that the semantics of all RG judgments, including the judgments
% for transaction-bound terms, make similar demands of the guarantee
% relation. Given that transactions are atomic (though not isolated), it
% is not immediately apparent why a transaction's guarantee is required
% to make explicit every step of its reduction. This requirement is
% justified however because, in reality, a transaction's atomicity is
% predicated on the isolation settings of the observer. A \iso{Read
%   Uncommitted} transaction, for example, is permitted to observe the
% internal state of a transaction $T$ even if $T$ is claimed to execute
% atomically.  In the interest of modular verification, the transaction
% must therefore make its internal state available via its guarantee
% relation.

% \begin{theorem}[\bfseries Soundness] 
% The rely-guarantee judgments defined by the rules in
% Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
% Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
% is provided in the supplementary material.}
% \end{theorem}

% \noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
% the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
% multi-step reduction of $c$ starting from a trace that satisfies $P$
% and $\I$, results in a trace that satisfies $Q$, and (b) the effect
% that every small-step of $c$ has on the trace is contained in $G$.
% Soundness of the RG judgment for transaction-bound commands
% ($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
% the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
% is always evaluated to a value that satisfies $C$.

>>>>

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference
may include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to the the inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not
extend naturally to isolation-constrained interference ($\R$)
because $R^{*}$ modulo $\I$ is not same as $\R^{*}$; the former is a
transitive relation constrained by $\I$, whereas the latter is the
transitive closure of a $\I$-constrained relation. We therefore
introduce a side-condition on $\I$ that restores the equality. The
condition requires $\I$ to allow an intereference $R^{*}(\stg,\stg'')$,
for two database states $\stg$ and $\stg''$, only if it also allows
interference for every prefix of $R^{*}(\stg,\stg'')$. In other words,
if $\I$ disallows interference from $\stg$ to $\stg'$, then an
$R$-step from $\stg'$ to $\stg''$ shouldn't make the interference from
$\stg$ to $\stg''$ valid. We call this the stability condition on
$\I$, defined as below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid, then an interference has already modified a
record, and no further interference will restore the original record,
because original record bears the id of a transaction that has long
committed. Thus, $\I_{ss}$ remains invalid.

@@

\subsection{Semantics and Soundness}

\begin{definition}[\bfseries Step-indexed reflexive transitive closure]
For all $A:\text{Type}$, $R: A \rightarrow A \rightarrow \mathbb{P}$, and $n :
\mathbb{N}$, the step-indexed reflexive transitive closure $R^n$ of $R$ is
the smallest relation satisfying the following
properties:
\begin{itemize}
\item $\forall (x:A).\, R^0 (x,x)$
\item $\forall (n:\mathbb{N})(x,y,z : A).\, R(x,y) \conj R^n(y,z) \Rightarrow
R^{n+1}(x,z)$
\end{itemize}
\end{definition}

\begin{definition}[\bfseries Interleaved step and multi-step relations]
An interleaved step relation interleaves thread-local reductions with
interference from concurrent threads captured as the rely relation
($R$).  It is defined thus:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{lcl}
\I \vdash (c,\E) \rstepsto (c',\E') & \defeq & \I \vdash 
  (c,\E) \stepsto (c',\E') \\
  &   & \disj (c' = c \conj R(\E, \E') \conj \I(\E'))\\
\end{array}
\end{smathpar}

\noindent The interleaved step relation for transaction bound expressions
($\txnbox{e}_i$) and commands ($\txnbox{c}_i$) is defined similarly.
An interleaved multi-step relation ($\stepssto{n}$) is the step-indexed
reflexive transitive closure of the interleaved step relation.
\end{definition}

\begin{definition}[\bfseries Semantics of the RG judgment]
\label{def:rg-semantics}
The semantics of the RG sextuple $\rg{\mathbb{I},P,R}{c}{G,Q}$ is defined
in terms of the interleaved step relation thus:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{l}
\hspace*{-0.3in}
\rg{\mathbb{I},P,R}{c}{G,Q} \;\defeq\; \forall \E.\, P(\E)
  \wedge \mathbb{I}(\E) \\
\hspace*{0.4in}\Rightarrow (\forall n,\E'.\; \I \vdash (c,\E) 
    \rstepssto{n} (\cskip,\E') \Rightarrow Q(\E')) \\
\hspace*{0.5in}\conj \texttt{step-guaranteed}(\I,R,G,c,\E)\\
\end{array}
\end{smathpar}

\noindent The first conjunct in the consequent is called the \emph{Hoare
consequent} since it ascribes Hoare triple semantics to an RG sextuple.
The second conjunct, called the \emph{guarantee consequent}, uses the
$\texttt{step-guaranteed}$ predicate defined below:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{l}
\texttt{step-guaranteed}(\I,R,G,c,\E) \;\defeq\; \forall n,\E',c'',\E''.\\
\hspace*{0.2in}\I \vdash (c,\E) \rstepssto{n} (c',\E') \conj \I \vdash (c',\E') \stepsto
  (c'',\E'') \Rightarrow G(\E',\E'')\\
\end{array}
\end{smathpar}

\noindent The guarantee consequent requires $G$ to capture the trace effect of
every small-step of $c$, where the reduction can be interleaved by the
interference ($R$) from concurrent threads. The semantics of the RG
sextuple for transaction-bound commands ($\txnbox{c}_i$) is defined
similarly. Expressions, unlike commands, evaluate to a value $v$, and
the semantics of their RG septuple ($\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$) differs slightly in that its
Hoare consequent requires the value $v$ to satisfy the assertion $C$. 
\end{definition}

Note that the semantics of all RG judgments, including the judgments
for transaction-bound terms, make similar demands of the guarantee
relation. Given that transactions are atomic (though not isolated), it
is not immediately apparent why a transaction's guarantee is required
to make explicit every step of its reduction. This requirement is
justified however because, in reality, a transaction's atomicity is
predicated on the isolation settings of the observer. A \iso{Read
  Uncommitted} transaction, for example, is permitted to observe the
internal state of a transaction $T$ even if $T$ is claimed to execute
atomically.  In the interest of modular verification, the transaction
must therefore make its internal state available via its guarantee
relation.

\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to the semantics of
Definition~\ref{def:rg-semantics}.\footnote{Formal proof of soundness
is provided in the supplementary material.}
\end{theorem}

\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b) the effect
that every small-step of $c$ has on the trace is contained in $G$.
Soundness of the RG judgment for transaction-bound commands
($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
is always evaluated to a value that satisfies $C$.
